---
title: "Project 1- Hockey- Colleen Moore"
date: "9/14/2020"
output:
  rmarkdown::github_document:
    toc: TRUE
    toc_depth: 3
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required Packages
The required packages to run the functions in this project are`jsonlite`,`httr`, `tidyverse`, and `knitr` . 
The `jsonlite` and `httr` packages are used in accessing the api's. `tidyverse` is used for data analysis and `knitr` is used for formatting. The tables returned from `kable`, a `knitr` function, are easier to read and print nicely. 
```{r, include= FALSE, warning= FALSE}
library(tidyverse)
library(jsonlite)
library(httr)
library(knitr)

```

# Write Functions to contact the [NHL records API](https://gitlab.com/dword4/nhlapi/-/blob/master/records-api.md) 
The five functions below to acess various parts of the the NHL records api. Data is returned in a data frame format for ease of analysis. 

## `franchise` function
Returns the franchise information.
The columns returned by this function include: <br>

* Id
* First season ID
* Last season ID
* Most recent team ID
* Team common name
* Team place name
* Total

The information returned is in the format of a data table. 

```{r, warning= F}
franchise <- function(fran){
  url<- "https://records.nhl.com/site/api/franchise"
  fran_get<- content(GET(url), "text")
  fran_json<- fromJSON(fran_get, flatten= T, simplifyDataFrame = T)
  fran_df<- as.data.frame(fran_json)
  colnames(fran_df) <- str_remove(colnames(fran_df), "data.")
  return(fran_df)
}

```

```{r, warning= F, message=F, include=F}
franchise()
```

## `fran_team_totals` function
Accesses the franchise team totals
This function returns Total stats for every franchise (ex roadTies, roadWins, etc)

```{r, warning= F, message= F}
fran_team_totals<- function(fran_team) {
  url<- "https://records.nhl.com/site/api/franchise-team-totals"
  franTT_get<- content(GET(url), "text")
  franTT_json<- fromJSON(franTT_get, flatten= T, simplifyDataFrame = T)
  franTT_df<- as.data.frame(franTT_json)
  colnames(franTT_df) <- str_remove(colnames(franTT_df), "data.")
  return(franTT_df)
}
  
```

```{r, message=F, include=F}
fran_team_totals()
```

## `team_data` function
Accesses team specific season records by team number or by common team name. 
In order to be able to acces the team common name I had to refer to my previous function above, the franchise function, that links up the team common name and the franchise id.
This function returns information on a specific team such as:

* Fewest losses
* Fewest wins
* Most game goals
* Winning Streak
* as well as a whole lot more information

```{r, message= F}

team_data<- function(team){
  if(is.character(team)){
    team_no<- franchise() %>% select(id, teamCommonName)%>% filter(teamCommonName== team)%>% select(id)
    base_url<- "https://records.nhl.com/site/api/franchise-season-records?cayenneExp=franchiseId="
    url<- paste0(base_url,team_no)
    team_get<- content(GET(url), "text")
    team_json<- fromJSON(team_get, flatten= T, simplifyDataFrame = T)
    team_df<- as.data.frame(team_json)
    colnames(team_df) <- str_remove(colnames(team_df), "data.")
  return(team_df)
  }
  if(is.numeric(team)){
  base_url<- "https://records.nhl.com/site/api/franchise-season-records?cayenneExp=franchiseId="
  url<- paste0(base_url,team)
  team_get<- content(GET(url), "text")
  team_json<- fromJSON(team_get, flatten= T, simplifyDataFrame = T)
  team_df<- as.data.frame(team_json)
  colnames(team_df) <- str_remove(colnames(team_df), "data.")
  return(team_df)
  }
}
```



## `Goalie_data` function
Accesses goalie records for the specified team entering either the franchise ID number or the common team name.
The function appears to return statistic information on each individual goalie that played for a specific franchise. It returns information such as: <br>

* most goals saved in one game
* number of seasons played
* most shout out in one season
* etc



```{r}
Goalie_data<- function(team){
  if(is.character(team)){
   team_no<- franchise() %>% select(id, teamCommonName)%>% filter(teamCommonName== team)%>% select(id)
    base_url<- "https://records.nhl.com/site/api/franchise-goalie-records?cayenneExp=franchiseId="
    url<- paste0(base_url, team_no)
    Gteam_get<- content(GET(url), "text")
    Gteam_json<- fromJSON(Gteam_get, flatten= T, simplifyDataFrame = T)
    Gteam_df<- as.data.frame(Gteam_json)
    colnames(Gteam_df) <- str_remove(colnames(Gteam_df), "data.")
  return(Gteam_df)
  }
  if(is.numeric(team)){
  base_url<- "https://records.nhl.com/site/api/franchise-goalie-records?cayenneExp=franchiseId="
  url<- paste0(base_url,team)
  Gteam_get<- content(GET(url), "text")
  Gteam_json<- fromJSON(Gteam_get, flatten= T, simplifyDataFrame = T)
  Gteam_df<- as.data.frame(Gteam_json)
  colnames(Gteam_df) <- str_remove(colnames(Gteam_df), "data.")
  return(Gteam_df)
  }
}
```



## `Skater_data` function
Accesses skater records for the specified team entering either the franchise ID number or the common team name. 
This function returns information such as:

* Players names
* Number of games played
* Numer of goals made
* Number of assists
* Number of seasons played
* etc


```{r}
Skater_data<- function(team){
  if(is.character(team)){
    team_no<- franchise() %>% select(id, teamCommonName)%>% filter(teamCommonName== team)%>% select(id)
    base_url<- "https://records.nhl.com/site/api/franchise-skater-records?cayenneExp=franchiseId="
    url<- paste0(base_url, team_no)
    Steam_get<- content(GET(url), "text")
    Steam_json<- fromJSON(Steam_get, flatten= T, simplifyDataFrame = T)
    Steam_df<- as.data.frame(Steam_json)
    colnames(Steam_df) <- str_remove(colnames(Steam_df), "data.")
  return(Steam_df)
  }
  if(is.numeric(team)){
  base_url<- "https://records.nhl.com/site/api/franchise-skater-records?cayenneExp=franchiseId="
  url<- paste0(base_url,team)
  Steam_get<- content(GET(url), "text")
  Steam_json<- fromJSON(Steam_get, flatten= T, simplifyDataFrame = T)
  Steam_df<- as.data.frame(Steam_json)
  colnames(Steam_df) <- str_remove(colnames(Steam_df), "data.")
  return(Steam_df)
  }
}
```

```{r, include=F}
Skater_data("Hurricanes")
```


# Access the stats API 
Write a function to access the [NHL stats API](https://gitlab.com/dword4/nhlapi/-/blob/master/stats-api.md)
Be able to get information corresponding to:

1. **active team roster for a specific team** <br>
  sepcify: stat_api= "roster" and a team number or team common name
2. **active team roster for a specific team with less information**<br>
  specify: stat_api= "less_roster" and a team number or team common name
3. **details of upcoming game for a team**<br>
  specify: stat_api= "schedule" and a team number or team common name
4. **details of last game played** <br>
  specify: stat_api= "prev_sch" and a team number or team common name
5. **Team stats for the season** <br>
  specify: stat_api= "stats" and a team number or team common name 
6. **Team roster for a season** <br>
  specify: stat_api= "season", a team number or team common name, and the start of the season year of interest (year1= )
7. **Multiple team Stats**<br>
  specify: stat_api= "mult_teams"
  This function allows you to compare up to 5 teams by team number. 
8. **Specify which stats to get**<br>
  specify: stat_api= "specific_stats", for example, if I was interested in Single season playoffs I could enter stat= SingleSeasonPlayoffs.The documentation for the function does not include what specific stats you can get.  

Notes: Have to enter the stat_api in quotes... so stat_api= "roster"

Make a table with the ID, common team name, and franchise ID to reference in functions. The stat API works on team number while the NHL records API works on franchise numbers. 
```{r, message= F}
team_url<- "https://statsapi.web.nhl.com/api/v1/teams/"
team_get<- content(GET(team_url), "text")
team_JSON<- fromJSON(team_get, flatten= T, simplifyDataFrame = T)
team_df<- as.data.frame(team_JSON)
colnames(team_df) <- str_remove(colnames(team_df), "teams.")
reference_table<- team_df %>% select(id, teamName, franchiseId)
```

Function allows you to specify either the team number or the teams common name. 
```{r}
stats_data<- function(team, stat_api, year1=NULL, stat= NULL, team2= NULL, team3= NULL, team4= NULL, team5= NULL){
  if(is.character (team)){
  team<- reference_table %>% filter(teamName== team)%>% select(id)
  }
  if(stat_api== "roster"){
  base_url<- "https://statsapi.web.nhl.com/api/v1/teams/"
  url<- paste0(base_url, team,"?expand=team.roster" )
  stat_team_get<- content(GET(url), "text")
  stat_team_json<- fromJSON(stat_team_get, flatten= T, simplifyDataFrame = T)
  stat_team_df<- as.data.frame(stat_team_json)
  return(stat_team_df$teams.roster.roster[[1]])
  }
  if(stat_api== "less_roster"){
  base_url<- "https://statsapi.web.nhl.com/api/v1/teams/"
  url<- paste0(base_url, team,"?expand=person.names")
  stat_team_get<- content(GET(url), "text")
  stat_team_json<- fromJSON(stat_team_get, flatten= T, simplifyDataFrame = T)
  stat_team_df<- as.data.frame(stat_team_json)
  return(stat_team_df)
  }
  if(stat_api== "schedule"){
  base_url<- "https://statsapi.web.nhl.com/api/v1/teams/"
  url<- paste0(base_url, team,"?expand=team.schedule.next")
  stat_team_get<- content(GET(url), "text")
  stat_team_json<- fromJSON(stat_team_get, flatten= T, simplifyDataFrame = T)
  stat_team_df<- as.data.frame(stat_team_json)
  return(stat_team_df)
  }
  if(stat_api== "prev_sch"){
  base_url<- "https://statsapi.web.nhl.com/api/v1/teams/"
  url<- paste0(base_url, team,"?expand=team.schedule.previous")
  stat_team_get<- content(GET(url), "text")
  stat_team_json<- fromJSON(stat_team_get, flatten= T, simplifyDataFrame = T)
  stat_team_df<- as.data.frame(stat_team_json)
  return(stat_team_df$teams.previousGameSchedule.dates[[1]][[6]][[1]])
  }
  if(stat_api== "stats"){
  base_url<- "https://statsapi.web.nhl.com/api/v1/teams/"
  url<- paste0(base_url, team,"?expand=team.stats")
  stat_team_get<- content(GET(url), "text")
  stat_team_json<- fromJSON(stat_team_get, flatten= T, simplifyDataFrame = T)
  stat_team_df<- as.data.frame(stat_team_json)
  return((stat_team_df[[9]][[1]])[[1]][[1]])
  }
  if(stat_api== "season"){
  year2<- year1 + 1
  base_url<- "https://statsapi.web.nhl.com/api/v1/teams/"
  url<- paste0(base_url, team,"?expand=team.roster&season=", year1,year2)
  stat_team_get<- content(GET(url), "text")
  stat_team_json<- fromJSON(stat_team_get, flatten= T, simplifyDataFrame = T)
  stat_team_df<- as.data.frame(stat_team_json)
  return(stat_team_df$teams.roster.roster[[1]])
  }
  if(stat_api== "mult_teams"){
  base_url<- "https://statsapi.web.nhl.com/api/v1/teams/"
  url<- paste0(base_url,"?teamId=", team,",", team2,",", team3,",", team4,",", team5)
  stat_team_get<- content(GET(url), "text")
  stat_team_json<- fromJSON(stat_team_get, flatten= T, simplifyDataFrame = T)
  stat_team_df<- as.data.frame(stat_team_json)
  return(stat_team_df)
  }
  if(stat_api== "specific_stats"){
  base_url<- "https://statsapi.web.nhl.com/api/v1/teams/"
  url<- paste0(base_url,team, "?stats=stats", stat)
  stat_team_get<- content(GET(url), "text")
  stat_team_json<- fromJSON(stat_team_get, flatten= T, simplifyDataFrame = T)
  stat_team_df<- as.data.frame(stat_team_json)
  return(stat_team_df)
  }
}



```

I tested each function above and each one seems to work fine. 

```{r, warning=F, include=F}
stats_data(stat_api = "roster", team= "Sabres")
```


```{r, warning=F, message= F, include=T}
stats_data(stat_api= "less_roster", team= "Hurricanes")
```

```{r, warning=F, message= F, include=F}
a<- stats_data(stat_api= "schedule", team= 54)
```

```{r, warning=F, message= F, include=F}
stats_data(stat_api= "prev_sch", team= 12)
```

```{r, warning=F, message= F, include=F}
stats_data(stat_api= "stats", team= "Blackhawks")
```

```{r, warning=F, message= F, include=F}
stats_data(stat_api= "stats", team= "Maple Leafs")
```

```{r, warning=F, message= F, include=F}
 stats_data(stat_api= "season", team= "Hurricanes", year1 = 2005)
```


```{r, warning=F, message= F, include=F}
stats_data(stat_api="mult_teams", team= 54, team2= 4, team3= 9, team4= 12)
```

```{r, warning=F, message= F, include=F}
stats_data(stat_api="specific_stats", team = "Hurricanes", stat= "SingleSeasonPlayoffs")
```

# Wrapper Function
This function is able to call any of the functions above in this project. The function name as well as the arguments needed for the function have to be supplied. So for example if you wanted goalie data for the Carolina Hurricanes, you would have to enter: <br>
NHL_master(func= "Goalie_data", team= "Hurricanes")

```{r, warning=F, message= F}
NHL_master<- function(func, team, stat_api= NULL, year1=NULL, stat= NULL, team2= NULL, team3= NULL, team4= NULL, team5= NULL,.....){
  if(func == "franchise"){
    return(franchise(team))
  }
  if(func== "fran_team_totals"){
  return(fran_team_totals(team))
  }
  if(func== "team_data"){
  return(team_data(team))
  }
  if(func== "Goalie_data"){
  return(Goalie_data(team))
  }
  if(func== "Skater_data"){
  return(Skater_data(team))
  }
  if(func=="stats_data"){
  return(stats_data(team, stat_api, year1, stat, team2, team3, team4, team5))
  }
}

```

I tried out my `NHL_master` function a couple of times and it appears to work fine. 

```{r, include= F}
NHL_master(func= "stats_data", stat_api= "mult_teams", team= 54, team2= 4, team3= 12)

NHL_master(func= "stats_data", stat_api= "season", team= "Hurricanes", year1= 2015)
```


# Data Exploration

## Is it true that teams tend to win more at home compared to on the road? 
I don't know too much about hockey, or any sport for that matter, but I have always heard about the home court advantage. Let's see if this dataset suppoorts that. The NHL records api contains information about the number of wins and losses at home and away. This can be accessed through the `fran_team_totals` function. There are 2 games types listed for most of the teams type 2 and type 3. Type 2 games are normal season games. I wil limit my analysis to type 2 games. 

```{r, message= F}

wins_losses<- fran_team_totals()%>% filter(gameTypeId == 2)%>% mutate(perc_wins= wins/gamesPlayed)

```
Looking at the wins_losess table, some teams have played more games than others. I assume this table has information since the team was formed and some teams have existed longer than others and thus have played more games. 

Normalize the number of wins and losses for each team by dividing the wins & losses by number of games played. Display the first 10 rows of the data frame. 
```{r, message=F}
perc_wins_losses<-wins_losses %>% mutate(perc_wins_home=homeWins/wins, perc_wins_road= roadWins/wins, perc_loss_home= homeLosses/losses, perc_loss_road= roadLosses/losses, perc_OT_loss_home=homeOvertimeLosses/overtimeLosses, perc_OT_loss_road= roadOvertimeLosses/overtimeLosses)%>% select(teamName, perc_wins_home,perc_wins_road, perc_loss_home, perc_loss_road, perc_OT_loss_home, perc_OT_loss_road) %>% arrange(desc(perc_wins_home))
kable(head(perc_wins_losses, 10), digits=2)
```

From the table above it looks like more than half of the percent wins are at home. Calculate the mean home vs. road wins. 
```{r}
a<- perc_wins_losses %>% summarize(meanWinsHome= mean(perc_wins_home), meanWinsRoad= mean(perc_wins_road), meanLossHome= mean(perc_loss_home), meanLossRoad= mean(perc_loss_road))
a
```
This tables show that teams tend to win more at home compared to on the road. 

## Are more aggressive teams (meaning thoes with higher pentalty minutes) more likely than average to win games? 
```{r, warning= F, message= F}
pentalty_wins<- wins_losses %>% mutate(perc_wins= wins/gamesPlayed, pent_per_game= penaltyMinutes/gamesPlayed) %>% select(teamName, perc_wins, pent_per_game)
avg_wins<- mean(pentalty_wins$perc_wins)
avg_pent<- mean(pentalty_wins$pent_per_game)
avg_wins
avg_pent


```
The average percentage of wins is about 40% of games played and teams have, on average, 13.05 minutes in the pentalty box. 


Since we know the aveage percentage of wins is about 40%, we can create a categorial variable for each team if they have greater than the average percentage of wins or less than the average percentage of wins. The same thing can be done for penatly minutes. 

```{r, message= F}
pentalty_wins$wins<- ifelse(pentalty_wins$perc_wins< avg_wins, "Below Avg Wins", "Above Avg Wins")
pentalty_wins$pent <- ifelse(pentalty_wins$pent_per_game < avg_pent, "Below Avg Pentalty Mins", "Above Avg Pentalty Mins")
```

```{r}
kable(table(pentalty_wins$wins, pentalty_wins$pent), caption= "Average wins vs. Average Pentalty Mins")
```

From the contigency table above it looks teams with above averge pentalty minutes tend to win more games compared to teams with below average pentalty minutes. 

## How many seasons do the Carolina Hurricanes keep their players?
```{r, message=F}
sk_hurricanes<- Skater_data("Hurricanes")

kable(table(sk_hurricanes$positionCode, sk_hurricanes$seasons), caption= "Number of seasons played by position")
```
The Carolina Hurricanes has a center for 16 seasons and a defenseman for 13 seasons, but in general most players, regarless of position, only stay for a couple of seasons. 

Histogram of # of seasons for each player type
```{r, message= F}
hist_player<- Skater_data("Hurricanes")
ggplot(hist_player, aes(seasons))+ geom_histogram(fill= "blue")+ labs(title= "Number of Seasons a player plays for the Carolina Hurricanes", x= "Number of Seasons Played" )
```
This histogram shows that most players only play on the Carolina Hurricanes for less than 4 seasons. 

## What positions spend the most time in the penalty box for the Caroliana Hurricanes? 
```{r, message= F}
ros<- NHL_master(func= "stats_data", stat_api = "roster", team= "Hurricanes")
sk<- NHL_master(func= "Skater_data", team= "Hurricanes")
```

Combine these 2 tables based on Player ID. Do an inner_join which will only keep players that are in both tables. 
```{r}
player_tbl<- sk %>% inner_join(ros, by= c("playerId" = "person.id" ))
```

```{r, message=F, warning= F}
graph<- player_tbl %>% group_by(position.name)%>% summarise(penalty_mins= mean(mostPenaltyMinutesOneSeason))
kable(graph, digits= 2)
```




Right wing players have the most penalty mins for the Carolina Hurricanes. Lets see the same data in a plot format. 

```{r}
ggplot(graph, aes(position.name, penalty_mins))+ geom_col(fill= "red")+ labs(x= "Position Name", y= "Most Penalty Minutes in One Season", title= "Which team positions are in the pentalty box the most for the Carolina Hurricanes?")
```



This plot shows that the right wing position is in the pentalty box the most for the Hurricanes. 


## Are teams that have played more games more likely to win? 

```{r}
wins_losses<- wins_losses%>% mutate(Percentage_wins= ifelse(perc_wins < avg_wins, "Below avg wins", "Above avg wins" )) %>% group_by(Percentage_wins)
wins_losses$Percentage_wins<- as.factor(wins_losses$Percentage_wins)
ggplot(wins_losses, aes(x= gamesPlayed, y= perc_wins, color= Percentage_wins)) + geom_point() + labs(title= "Are experienced teams more likely to win more games?", x= "Number of Games Played", y= "Percentage of Wins")
```

This plot shows that the teams that have played more games, tend to have a higher winning percentage but it levels off around the 2000th game. The teams in blue are the teams will a less than average winning percentage and thoes in red have a higher than average winning percentage.  

## Is there a relationship between number of goals scored and number of assists for the Carolina Hurricanes? Do players that score more goals tend to have less assists? 

```{r}
ggplot(hist_player, aes(mostGoalsOneSeason, mostAssistsOneSeason, color= positionCode))+ geom_point() + labs(title= "Goals vs Assists by Position", x= "Most Goals scored in One Season", y= "Most Assists in One Season")
```

While each players record number of goals and assists may not be in the same season, we can see a general trend that players that score more also tend to have a higher number of assists. Defenseman tend to have the lowest number of goals and assists, which I think makes sense since their main objective to do defend their goal. 

## Comparison of the Carolina Hurricanes and the Edmonton Oilers
In 2006 the Carolina Hurricanes won the Stanley Cup against the Edmonton Oilers. 
Let's explore these two teams. 
```{r}
standly_team_names<- NHL_master(func= "stats_data", stat_api= "season", team= "Hurricanes", year1= 2005)%>%select(person.id, person.fullName, position.name)
```

```{r, message= F}
Cup_Hurricanes_data<- standly_team_names %>% inner_join(Skater_data("Hurricanes"), by= c("person.id"= "playerId"))
```

```{r, message=F}
Oilers_names<- NHL_master(func= "stats_data", stat_api= "season", team= "Oilers", year1= 2005)%>% select(person.id, person.fullName, position.name) 
```

```{r, message=F}
Oilers_data<- Oilers_names %>% inner_join(Skater_data("Oilers"), by= c("person.id"= "playerId"))

```

The players that were on the Hurricanes and Oilers team during the 2005- 2006 season was obtained using the `NHL_master` function. These tables were then used to get player specific information from the `Skater_data` function. These two tables were combined to form the Standly_team_2005 table, which was then used for futher analysis. 


```{r, message=F}
Standly_team_2005<- rbind(Cup_Hurricanes_data, Oilers_data)
```

```{r}
ggplot(Standly_team_2005,aes(x= position.name, y= goals, fill= franchiseName)) + geom_boxplot() + labs(x= "Position Name", y= "Overall number of goals for each player", title= "Comparison of the 2006 Standley Cup Teams: Goals")
```


The centers and defenseman on the Carolina Hurricanes 2005-2006 team had a higher median number of goals compared to the Oilers. However the Oilers right wing players had a higher median number of goals compared to the right wings on the Hurricanes.  

```{r}
ggplot(Standly_team_2005,aes(x= position.name, y= gamesPlayed, fill= franchiseName)) + geom_boxplot() + labs(x= "Position Name", y= "Overall number of games played", title= "Comparison of the 2006 Standley Cup Teams: Number of Games Played")
```

The Hurricanes had more experiences centers and defenseman compared to the Oilers. While the Oilers had more experienced right wing and left wing players. 

```{r}
ggplot(Standly_team_2005,aes(x= position.name, y= mostPenaltyMinutesOneSeason, fill= franchiseName)) + geom_boxplot() + labs(x= "Position Name", y= "Most Penalty Minutes in One Season", title= "Comparison of the 2006 Standley Cup Teams: Penalties")
```

The thing that stands out to me in this boxplot is that the left wing players on the Oilers had a much higher median of penalty minutes in one season. 





